tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-14 paper.tex
Tweaks to intro

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
Despite these successes and the ongoing collection of population-scale
Biobank data~\citep{tanjo2021practical}, there are substantial open
questions around GWAS methology.
% TODO rephrase
In particular, correctly accounting for population stratification
remains challenging~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these breakthroughs, substantial progress has been made
on statistical genetics questions.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,baumdicker2022efficient,anderson2023}
have been instrumental in making simulations of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store vast genetic variation datasets very compactly. Rather than
storing the genetic sequence variation in its conventional matrix
form, we can instead store the underlying genealogical history
and mutations that ultimately \emph{result in} the observed
variation~\citep{kelleher2019}.
For example, the ultra-realistic simulations of 1.4 million
French-Canadian individuals provided by \citet{anderson2023}
requires around 550 GB of storage space in the
standard gzip-compressed VCF format ($\sim$280TB uncompressed)
for chromosome 9. The original simulation file representing
precisely the same genetic variation data, \emph{plus} the
complete simulated ARG, requires only 1.36GB of space
in the compressed \texttt{tszip} format. Thus, in this
case we achieve more than 400-fold compression,
and this massive and otherwise highly unwieldy dataset
becomes straightforward to work with (we discuss
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

@@

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG.
The package is build on top of
\texttt{tskit}~\citep{ralph2020}, which provides a powerful
framework for storing and processing ARGs.
Using these efficient data structures, \texttt{tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.

>>>>

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
This success has led to the collection of many population-scale Biobank
datasets~\citep{tanjo2021practical}.
This dense sampling and resulting vast volumes of data, however,
are presenting significant challenges to GWAS methology.
For example, how to correctly account for population stratification
remains an open question~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs in inference have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these recent developments, substantial progress has been made
on statistical genetics applications.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store vast genetic variation datasets very compactly. Instead of
storing the genetic sequence variation in the conventional
matrix-based form, we can store the underlying genealogical history
and mutations that ultimately \emph{result} in the observed
variation~\citep{kelleher2019}.
For example, a highly realistic simulation
of chromosome 9 for 1.4 million
French-Canadian individuals~\citep{anderson2023}
requires around 550GB of storage space in the
standard gzip-compressed VCF~\citep{danecek2011}
format ($\sim$280TB uncompressed).
The original simulation file representing
precisely the same genetic variation data, plus the
complete simulated ARG, requires only 1.36GB of space
in the compressed \texttt{tszip} format. Thus,
by encoding in terms of the underlying ARG,
in this case we achieve more than 400-fold compression,
and this highly unwieldy dataset
becomes straightforward to work with (we discuss
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

@@

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
Using the efficient data structures provided by \texttt{tskit},
\texttt{tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.
